Eagle: multi-locus association mapping on a genome-wide scale made routine

### 2019-08-02 BioinformaticsRevison/bioinformaticspaper.tex
Review 1 revisions almost complete - waiting on analyses

\begin{figure}
\caption{
\setstretch{1.6}{
Power verse false positive rates for Eagle and the single-locus methods GEMMA and FaST-LMM. 
FaST-LMM was run where all the SNP data are used 
to estimate the relationship matrix (FaST-LMM$^{all}$)   and where genotype data from every five-hundredth SNP are used to 
estimate the relationship matrix (FaST-LMM$^{few}$). Eagle has substantially higher power than the single-locus methods.   }
}
\vspace{-3cm}
\label{figpowersingle}
%%\begin{center}
\includegraphics[width=15cm, height=20cm]{power2main.eps}
%%\end{center}

@@

We repeated the single-locus analyses as first performed \citep{nicod2016genome} but with a few exceptions. We 
focused on autosomal SNPs, our analyses were based on SNP genotypes  rather than SNP dosages, 
we sought to control the false positive rate not false discovery rate of the methods, and significance thresholds were 
found empirically via permutation \citep{doerge1996permutation}.

We ran Eagle in three ways. 
Eagle chooses the best model via the extended Bayesian information criteria (extBIC) \citep{chen2008extended}. 
  The conservativeness of the extBIC can be adjusted by a single regularisation parameter  $\gamma$ that ranges from zero to one. 
In the simulation study, this parameter was set to one, its most conservative and default setting. The mouse data were also analysed 
under this setting (Eagle$^{default}$).  An alternate  \citep{chen2008extended} , less conservative way of setting $\gamma$ is to  let $\gamma = 1 - \frac{1}{(2\kappa)}$ with $\kappa=\frac{log(L)}{log(n_g)}$ where 
 $L$ is the number of loci that span the genome, and $n_g$ is the number of individuals/groups/lines/strains in the study
  (Eagle$^{alt}$). However, our preferred way is to set the $\gamma$ parameter for each trait via permutation (Eagle$^{optimal}$).  We used 100 permutations to set $\gamma$ to give a false  positive rate of 5\%.  This only took six times as long as a single analysis of the data. This is because the marker data need only be read  once, 
 and only the trait data changes across permutations leading to other computational efficiencies.  
  This permutation method has been implemented within the
 Eagle package.

The genome wide results from the analyses of the mouse data are shown in Figure  \ref{figmouse}. The mouse study recorded
measurements on 200 traits. Of these, in the original study, 45 were able to have their findings  corroborated by previously published work. We focused 
our analyses here on these same 45 traits. Overall,  FaST-LMM$^{Bonf}$,  FaST-LMM$^{perm}$, Eagle$^{default}$, Eagle$^{alt}$, and Eagle$^{optimal}$ found 50, 37, 67, and 106, SNP-trait findings, respectively, across 39 traits. No associations were found by FaST-LMM and Eagle for the other six traits. 
Eagle$^{alt}$ and Eagle$^{optimal}$ also found SNP-trait associations not found in the original study. This is despite their analyses being based 
on the SNP genotype data and the original study being based on SNP dosage data. Eagle$^{alt}$ found two  and 
Eagle$^{optimal}$ found seven new findings (Supplementary Table \ref{suptabnew}).  These new findings all involved SNPs whose association had been confirmed for other related traits in the original 
study.

In the simulation study, Eagle outperforms single-locus association mapping. Here, Eagle$^{default}$, where $\gamma=1$, finds less associations 
than FaST-LMM. Why the discrepancy in performance?   The answer lies in the conservativeness of Eagle.  With the added genetic complexity implicit within the mouse data, Eagle is more conservative when $\gamma$ is set to one than in the simulation study.  However, the relative results of the simulation study remain true. For similar false positve rates, Eagle is superior to single-locus association mapping. As a case in point, here FaST-LMM found 50 SNP-trait associations with a false positive rate of 5\%. Eagle, with the same false positive rate (Eagle$^{optimal}$) found 106 SNP-trait associations, more 
than a 100\% increase in findings.

\begin{figure}
\caption{
\setstretch{1.1}{
Genome-wide association mapping results from analyses of the mouse data for the single-locus method FaST-LMM and the 
multi-locus method Eagle. Genome-wide significance thresholds for FaST-LMM were calculated via the Bonferroni correction 
(FaST-LMM$^{Bonf}$) and permutation (FaST-LMM$^{perm}$). 
Eagle was run under three settings; its default setting (Eagle$^{default}$),  an alternate less conservative setting based on 
the number of SNPs and sample size (Eagle$^{alt}$), and where the model selection had been optimised for a false positive rate 
of 5\% (Eagle$^{optimal}$).   The number of SNP-trait associations found are reported in the cells. }
}
\label{figmouse}
\begin{center}
%% obtained from bracewell by running ResultsV2.R from /data/geo047/MWAM/Mouse/OutBred/RScripts/
\includegraphics[width=12cm, height=12cm]{mouseresults.eps}
\end{center}

@@

\section{Discussion/Conclusion}
Eagle is a new linear mixed model based method (and R package) for multi-locus association mapping. It advances the state of association mapping in several ways. 
First, its computational footprint is much smaller than other multi-locus implementations. Eagle makes multi-locus analysis 
practical, even when the datasets are large. Second, the results from
 Eagle are immediately interpretable. They are the set of SNPs in strongest association with the trait where 
each SNP identifies a separate genomic region of interest. Third, it treats association mapping as a model selection problem, avoiding 
 multiple testing issues. 
 As we saw in the simulation study, Eagle has considerably higher power than single-locus methods but is comparable in run time.
Also, when analysing the mouse data, Eagle found more than double the SNP-trait associations than 
with single-locus association mapping, the method of choice. Furthermore, these extra findings were all true.

@@

\section*{Acknowledgements}
We would like to thank the IM\&T Scientific Computing group at the Commonwealth Scientific and Industrial Research Organisation (CSIRO) for providing access and support to the CSIRO Accelerator Cluster upon which all computation was performed.

>>>>

\begin{figure}
\caption{
\setstretch{1.6}{
Power verse false positive rates for Eagle and the single-locus methods GEMMA and FaST-LMM. 
FaST-LMM was run where all the SNP data are used 
to estimate the relationship matrix (FaST-LMM$^{all}$)   and where genotype data from every five-hundredth SNP are used to 
estimate the relationship matrix (FaST-LMM$^{few}$). Eagle has substantially higher power than the single-locus methods.   }
}
\vspace{-3cm}
\label{figpowersingle}
%%\begin{center}
\includegraphics[width=10cm, height=15cm]{power2main.eps}
%%\end{center}

@@

We repeated the single-locus analyses as first performed \citep{nicod2016genome} but some exceptions. We 
focused on autosomal SNPs, our analyses were based on SNP genotypes  rather than SNP dosages, 
we sought to control the false positive rate not false discovery rate of the methods, and significance thresholds were 
found empirically via permutation \citep{doerge1996permutation}.

We ran Eagle in three ways. 
Eagle chooses the best model via the extended Bayesian information criteria (extBIC) \citep{chen2008extended}. 
  The conservativeness of the extBIC can be adjusted by a single regularisation parameter  $\gamma$ that ranges from zero to one. 
In the simulation study, this parameter was set to one, its most conservative and default setting. The mouse data were also analysed 
under this setting (Eagle$^{default}$).  An alternate  \citep{chen2008extended}, less conservative way of setting $\gamma$ is to  let $\gamma = 1 - \frac{1}{(2\kappa)}$ with $\kappa=\frac{log(L)}{log(n_g)}$ where 
 $L$ is the number of loci that span the genome, and $n_g$ is the number of individuals/groups/lines/strains in the study
  (Eagle$^{alt}$). However, our preferred way is to set the $\gamma$ parameter for each trait via permutation (Eagle$^{optimal}$).  We used 100 permutations to set $\gamma$ to give a false  positive rate of 5\%.  This only took six times as long as a single analysis of the data. This is because the marker data need only be read  once, 
 and only the trait data changes across permutations leading to other computational efficiencies.  
  This permutation method has been implemented within the
 Eagle package.

The genome wide results from the analyses of the mouse data are shown in Figure  \ref{figmouse}, 
\textcolor{red}{
with the Manhattan plots of the 
single-locus analysis shown in Supplementary Figure \ref{supmanhattan}.} 
The mouse study recorded
measurements on 200 traits. Of these, in the original study, 45 were able to have their findings  corroborated by previously published work. We focused 
our analyses here on these same 45 traits. Overall,  FaST-LMM$^{Bonf}$,  FaST-LMM$^{perm}$, Eagle$^{default}$, Eagle$^{alt}$, and Eagle$^{optimal}$ found 47, 68, 37, 67, 106, SNP-trait findings, respectively, across 39 traits. No associations were found by FaST-LMM and Eagle for the other six traits. 
Eagle$^{alt}$ and Eagle$^{optimal}$ also found SNP-trait associations not found in the original study. This is despite their analyses being based 
on the SNP genotype data and the original study being based on SNP dosage data. Eagle$^{alt}$ found two  and 
Eagle$^{optimal}$ found seven new findings (Supplementary Table \ref{suptabnew}).  These new findings all involved SNPs whose association had been confirmed for other related traits in the original 
study.

In the simulation study, Eagle outperforms single-locus association mapping. Here, Eagle$^{default}$, where $\gamma=1$, finds less associations 
than FaST-LMM. Why the discrepancy in performance?   The answer lies in the conservativeness of Eagle.  With the added genetic complexity implicit within the mouse data, Eagle is more conservative when $\gamma$ is set to one than in the simulation study.  However, the relative results of the simulation study remain true. For similar false positve rates, Eagle is superior to single-locus association mapping. As a case in point, here FaST-LMM$^{perm}$ found 68 SNP-trait associations with a false positive rate of 5\%. Eagle, with the same false positive rate (Eagle$^{optimal}$) found 106 SNP-trait associations, more 
than a 50\% increase in findings.

\begin{figure}
\caption{
\setstretch{1.1}{
Genome-wide association mapping results from analyses of the mouse data for the single-locus method FaST-LMM and the 
multi-locus method Eagle. Genome-wide significance thresholds for FaST-LMM were calculated, via the Bonferroni correction 
(FaST-LMM$^{Bonf}$) and permutation (FaST-LMM$^{perm}$), at the 5\% significance level. 
Eagle was run under three settings; its default setting (Eagle$^{default}$),  an alternate less conservative setting based on 
the number of SNPs and sample size (Eagle$^{alt}$), and where the model selection had been optimised for a false positive rate 
of 5\% (Eagle$^{optimal}$).   The number of SNP-trait associations found are reported in the cells. }
}
\label{figmouse}
\begin{center}
%% obtained from bracewell by running ResultsV2.R from /data/geo047/MWAM/Mouse/OutBred/RScripts/
\includegraphics[width=12cm, height=12cm]{mouseresults.eps}
\end{center}

@@

\section{Discussion/Conclusion}
Eagle is a new linear mixed model based method (and R package) for multi-locus association mapping. It advances the state of association mapping in several ways. 
First, its computational footprint is much smaller than other multi-locus implementations. Eagle makes multi-locus analysis 
practical, even when the datasets are large. Second, the results from
 Eagle are immediately interpretable. They are the set of SNPs in strongest association with the trait where 
each SNP identifies a separate genomic region of interest. Third, it treats association mapping as a model selection problem, avoiding 
 multiple testing issues. 
 As we saw in the simulation study, Eagle has considerably higher power than single-locus methods but is comparable in run time.
Also, when analysing the mouse data, Eagle found more than double the SNP-trait associations than 
with single-locus association mapping, the method of choice. \textcolor{red}{Furthermore, because we converted the SNP dosages of the 
original study into genotypes and focusing our analyses on these data, the validity of the extra findings can be confirmed 
against the original findings}. These extra findings were all true.

@@

\section*{Acknowledgements}
We would like to thank the IM\&T Scientific Computing group at the Commonwealth Scientific and Industrial Research Organisation (CSIRO) for providing access and support to the CSIRO Accelerator Cluster upon which all computation was performed. 
We also gratefully acknowledge the comments of two anonymous reviewers whose suggestions have helped "polish" this paper.

### 2019-08-02 BioinformaticsRevison/supmaterials.tex
significant changes - FDR -> FPR. Revising heat map, adding Manhattan plots to sup materials, added Bonf correction to FaST-LMM etc
Review 1 revisions almost complete - waiting on analyses

\begin{figure}
\caption{Manhattan plot of single-locus analysis of the mouse data with FaST-LMM for the traits BMC Kurt,  BMC Kurt N, BMC Max, and BMC Max N}
\begin{center}
\includegraphics[width=15cm, height=15cm]{1.jpeg}
\end{center}
\end{figure}

\begin{figure}
\caption{Manhattan plot of single-locus analysis of the mouse data with FaST-LMM for the traits BMC Mean,  BMC Mean N, BMC Median, and BMC Perim}
\begin{center}
\includegraphics[width=15cm, height=15cm]{5.jpeg}
\end{center}
\end{figure}

>>>>

\begin{figure}
\caption{Manhattan plots of the single-locus analysis of the mouse data with FaST-LMM. 
Genome-wide significance thresholds for FaST-LMM were calculated, via the Bonferroni correction 
(pink horizontal line) and permutation (blue horizontal line), at a 5\% significance level. 
}
\label{supmanhattan}
\begin{center}
\includegraphics[width=15cm, height=15cm]{1.jpeg}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=15cm, height=15cm]{5.jpeg}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=15cm, height=15cm]{9.jpeg}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=15cm, height=15cm]{13.jpeg}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=15cm, height=15cm]{17.jpeg}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=15cm, height=15cm]{21.jpeg}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=15cm, height=15cm]{25.jpeg}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=15cm, height=15cm]{29.jpeg}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=15cm, height=15cm]{33.jpeg}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=15cm, height=15cm]{37.jpeg}
\end{center}
\end{figure}
